Evaluating `Graphical Perception' with CNNs

### 2018-08-15 PAPER/00_introduction.tex
Update paper from overleaf.

\firstsection{Introduction}

\maketitle

Convolutional neural networks (CNNs) have been successfully applied to a wide range of visual tasks, most famously to natural image object recognition~\cite{krizhevsky_imagenet2012, simonyan_very_deep2014, szegedy2015}, for which some claim equivalent or better than human performance. This performance comparison is often motivated by the idea that CNNs model or reproduce the early layers of the visual cortex, even though they do not incorporate many details of biological neural networks or model higher-level abstract or symbolic reasoning~\cite{yamins2016using, hassabis2017neuroscience, human_vs_machine_vision}. While CNN techniques were originally inspired by neuroscientific discoveries, recent advances in processing larger datasets with deeper networks have been the direct results of engineering efforts. Throughout this significant advancement, researchers have aimed to understand why and how CNNs produce such high performance~\cite{goodfellow_book, deeplearning_blackbox2017}, with recent works targeting the systematic evaluation of the visual perception limits of CNNs~\cite{clevr, not_so_clevr}.

One fundamental application of human vision is to understand data visualizations. This is a task unlike processing natural images, where real-world objects and their effects are abstracted into data and represented with visual marks. As a field, visualization catalogues and evaluates human perception of these marks, such as in the seminal \emph{graphical perception} experiments of Cleveland and McGill~\cite{cleveland_mcgill}. This work describes nine elementary perceptual reasoning tasks, such as position relative to a scale, length, angle, area, and shading density, plus orders their reasoning difficulty. But, with increasing research interest in the machine analysis of graphs, charts, and visual encodings, it seems pertinent to question whether CNNs are able to process these basic graphical elements and derive useful measurements from the building blocks of information visualization.

As such, we reproduce Cleveland and McGill's human perceptual experiments with CNNs, and discuss to what extent they have `graphical perception'. To perform this evaluation, we parametrize the elementary perceptual tasks and experiments suggested by Cleveland and McGill~\cite{cleveland_mcgill}, and define a set of regression tasks to estimate continuous variables. Against human perception, we pit four neural networks: a three-layer multilayer perceptron (MLP), the LeNet 2-layer CNN~\cite{lenet}, the VGG 19-layer CNN~\cite{simonyan_very_deep2014}, and the Xception 36-layer CNN~\cite{xception}. As CNNs trained on natural images are said to mimic layers of the human visual cortex, we investigate whether using weights trained on natural images (via ImageNet~\cite{imagenet}) or weights trained from scratch on elementary graphical perception tasks produces more accurate measurements and greater generalization.

We test these four networks across four scenarios presented by Cleveland and McGill~\cite{cleveland_mcgill}: 1) Nine elementary perceptual tasks with increasing parameteric complexity, e.g., length estimation with fixed x, then with varying x, then with varying width, including cross-network evaluations testing the generalizability of networks to unseen parameters; 2) The position-angle experiment, which compares judgements of bar charts to pie charts, 3) The position-length experiment, which compares grouped and divided bar charts, and 4) The bars and framed rectangles experiments, where visual cues aid ratio perception. We also investigate whether our CNNs can detect a proportional change in a measurement across scales, in relation to Weber's law.

With these experiments, we describe a ranking defining the ease with which our tested CNN architectures can estimate elementary perceptual tasks, as an equivalent to Cleveland and McGill's ranking for human perception. Further, we discuss the implications of our results and derive recommendations for the use of CNNs in perceiving visualizations. We accompany this paper with open source code and our input and results data, both to enable reproduction studies and to spur new machine perception systems more adept at graphical perception: \url{http://rhoana.org/perception}

>>>>

\firstsection{Introduction}
\maketitle
\vspace{-0.1cm}

%One fundamental application of human vision is to understand data visualizations. This is a task unlike natural image processing but which includes the abstraction of real-world objects and their effects into data, represented with visual marks. As a field, visualization catalogues and evaluates human perception of these marks, such as in the seminal \emph{graphical perception} experiments of Cleveland and McGill~\cite{cleveland_mcgill}. This work describes nine elementary perceptual reasoning tasks, such as position relative to a scale, length, angle, area, and shading density, plus orders their reasoning difficulty.

Convolutional neural networks (CNNs) have been successfully applied to a wide range of visual tasks, most famously to natural image object recognition~\cite{simonyan_very_deep2014, szegedy2015}, for which some claim equivalent or better than human performance. This performance comparison is often motivated by the idea that CNNs model or reproduce the early layers of the human visual cortex, even though they do not incorporate many details of biological neural networks or model higher-level abstract or symbolic reasoning~\cite{yamins2016using, hassabis2017neuroscience, human_vs_machine_vision}. While CNN techniques were originally inspired by neuroscientific discoveries, recent advances in processing larger datasets with deeper networks have been the direct results of engineering efforts. Throughout this significant advancement, researchers have aimed to understand why and how CNNs produce such high performance~\cite{deeplearning_blackbox2017}, \change{with recent works targeting the systematic evaluation of the limits of feed-forward convolutional neural networks for both image recognition problems \cite{Azulay2018} and for visual relation problems~\cite{clevr, not_so_clevr}.}

In visualization, there is increasing research interest in the computational analysis of graphs, charts, and visual encodings~\cite{maneesh_deconstructing_d3,Pineo2012_computational_perception,kafle2018dvqa}, for applications like information extraction and classification, visual question answering (``computer, which category is greater?''), or even design analysis and generation~\cite{Viegas2016}. One might turn to a CNN for these tasks. However, computational analysis of visualizations is a more complex task than natural image classification~\cite{Kahou2018}, requiring the identification, estimation, and relation of visual marks to extract information. For instance, we take for granted the human ability to generalize an understanding of length to a previously unseen chart design, or to estimate the ratios between lengths, yet for a CNN these abilities are in question, with no clear mechanism for concept abstraction.

% this is a nicely readable motivation!
Our goal is to better understand the abilities of CNNs for visualization analysis, and so we investigate the performance of current off-the-shelf CNNs on visualization tasks and show what they can and cannot accomplish. As computational visualization analysis is predicated upon an understanding of elementary perceptual tasks, we consider the seminal \emph{graphical perception} settings of Cleveland and McGill~\cite{cleveland_mcgill}. This work describes nine reasoning tasks, such as position relative to a scale, length, angle, area, and shading density, and measures human graphical perception performance on bar and pie chart quantity estimation. We reproduce Cleveland and McGill's settings with four different neural network designs of increasing sophistication (MLP, LeNet, VGG, and Xception), and compare their performance to human graphical perception. For this task, we collect new human measures for each elementary task, for the bars and frames rectangles setting, and for a Weber's law point cloud experiment. Further, as CNNs trained on natural images are said to mimic early human vision, we investigate whether using pre-trained natural image weights (via ImageNet~\cite{imagenet}) or weights trained from scratch on elementary graphical perception tasks produces more accurate predictions.

%  and greater generalization

%To perform this evaluation, we parametrize the elementary perceptual tasks and experiments suggested by Cleveland and McGill~\cite{cleveland_mcgill}, and define a set of regression tasks to estimate continuous variables. 
%We pit four neural networks against human perception: a three-layer multilayer perceptron (MLP), the LeNet 2-layer CNN~\cite{lenet}, the VGG 16-layer CNN~\cite{simonyan_very_deep2014}, and the Xception 36-layer CNN~\cite{xception}.

%With these experiments, we describe a ranking defining the ease with which our tested CNN architectures can estimate elementary perceptual tasks, equivalent to Cleveland and McGill's ranking for human perception.

%We test these four networks across four scenarios presented by Cleveland and McGill~\cite{cleveland_mcgill}: 1) Nine elementary perceptual tasks with increasing parameteric complexity, e.g., length estimation with fixed x, then with varying x and y, then with varying width, including cross-network evaluations testing the generalizability of networks to unseen parameters; 2) The position-angle experiment, which compares judgments of bar charts to pie charts; 3) The position-length experiment, which compares grouped and divided bar charts; and 4) The bars and framed rectangles experiments, where visual cues aid ratio perception. We also investigate whether our CNNs can detect a proportional change in a measurement across scales in relation to Weber's law.

First, we find that CNNs can more accurately predict quantities than humans for nine elementary perceptual tasks, but only if their training data includes similar stimuli. Second, that our networks can estimate bar chart lengths and pie segment angles with accuracy similar to humans, and that our networks trained more easily on bar charts. Third, that our networks were largely unable to estimate length ratios across five bar chart designs, unlike humans, and that bar chart design type had largely no effect on CNN performance but a significant effect on human performance. Fourth, that framing bars make it no easier for our CNNs to estimate just noticeable differences in length, unlike for humans. Fifth, that some CNNs can solve a difficult Weber's law problem that is beyond human ability. Practically, we find that networks trained from scratch on visualizations perform better than using pre-trained natural images weights, and that current Xception networks perform worse than older VGG networks.

A second goal of our work is to help frame the visualization community's discussion of CNNs as it builds towards future computational visualization analysis applications. For this, our findings suggest that CNNs are not currently a good model for human graphical perception, and as such their application to specific visualization problems may be possible but needs care. We discuss this in more detail with respect to designing CNNs for analyzing visualizations. Further, toward this goal, we accompany this paper with open source code and data, both to enable reproduction studies and to spur new machine perception systems more adept at graphical perception: \url{http://vcglab.org/perception}

### 2018-08-15 PAPER/01_related_work.tex
Update paper from overleaf.

\section{Previous Work}

\noindent{\textbf{Graphical Perception.}} Cleveland and McGill~\cite{cleveland_mcgill,cleveland1985graphical} coin the phrase \emph{graphical perception} to describe how different visual attributes and encodings are perceived by humans. They define \emph{elementary perceptual tasks} as mental-visual stimuli to understand encodings in visualizations, and declare a ranking based on their perceptual difficulty. From these definitions, the authors propose and perform different experiments such as the \emph{position-angle} experiment which compares bar charts and pie charts, the \emph{position-length} experiment where users judge relations between encoded values in grouped and divided bar charts, and the \emph{bars-and-framed-rectangles} experiment to evaluate Weber's law \cite{harrison2014_webers_law_rank} using the proportional relation between an initial distribution density and perceivable change.

Heer and Bostock later reproduced the Cleveland-McGill experiments via crowd-sourcing on Mechanical Turk~\cite{HeerBostock2010}, with similar results. Harrison \textit{et al.}~\cite{harrison2013influencing} repeated the Cleveland-McGill experiments while observing viewer emotional states, again with similar results. Our experimental setup again reproduces the Cleveland-McGill experiments, but instead of judging human perception, we judge machine perception using convolutional neural networks. While we focus on Cleveland and McGill's work from the mid 1980s due to their repeated reproduction, many other works also investigate human perception to visual encoding~\cite{bertin1967semiologie,treisman1988feature,carpendale2003considering,wilkinson2006grammar,widgor_perception2007,munzner2015visualization}.

@@

%\textbf{Comparing Visual Encodings.} Different visual encodings have advantages or disadvantages and the community does a great job comparing them. Higher-level comparisons include 2D versus 3D vector field and rendering studies~\cite{mckenzie_2d_3d,forsberg2009comparing_3d_vector,laidlaw_2d_vector,borkin2011arteries}, timeseries~\cite{herr2009timeseries} and scatterplots~\cite{tremmel1995visual,Wang_linegraph_vs_scatterplot}. Lower-level experiments target - besides others - open versus closed encodings~\cite{open_vs_closed_shapes}, and several evaluations of color space~\cite{ware1988color,rheingans1992color,Rogowitz2001_colormaps,kindlmann2002color}. While we investigate lower-level visual encodings in this work, we delay colorspace experiments for future work.

\noindent{\textbf{Computational Visualization Understanding.}} Pineo \textit{et al.}~\cite{Pineo2012_computational_perception} create a computational model of human vision based on neural networks. Their simulations show that understanding visualization triggers neural activity in high-level areas of cognition, with the authors suspecting that this activity is supported by low-level neurons performing elementary perceptional tasks. Other work tries to parse infographics by finding higher-level saliency models~\cite{bylinskii2016should}, or by extracting text or key visual elements from visualizations~\cite{diagram_understanding,kembhavi2016diagram,zoya_text_visual_tags}. However, these works do not investigate computational understanding of elementary perceptual tasks such as curvature, lengths, or position, which are the building blocks of visualization.

@@

\noindent{\textbf{Visual-cortex-inspired Machine Learning.}} The human visual cortex allows us to recognize objects in the world seemingly without effort (though few remember their infancy). This visual system is organized in layers, which inspired computational classifiers based on multilayer neural networks. Fukushima and Miyake developed the early Neocognitron quantitative model~\cite{fukushima1982neocognitron}, which ultimately led to the work of Hinton, Bengio, and LeCun~\cite{lecun2015deep} and today's GPU-powered \emph{deep} neural networks. Such networks exist with many architectures. For this paper, we compare a set of networks with different architectures and depths, plus networks with weights trained on natural images and on the elementary perceptual reasoning tasks themselves.
%Object detection is extremely difficult and therefore is especially impressive as light intensities can change by levels of magnitude and contrast between foreground and background is so often low. 
%In addition, the visual scene changes every time the human body or human eyes move. 
%This visual system exhibits a very noisy structure but because it is organized by layers it has inspired the mathematical theory of multilayer neural networks.

@@

%The reported classification performance is superior to that of humans and the question in regards to their functionality opens an interesting research topic.
%Visual cortex inspired machine learning classifiers exist with many different architectures. We select the traditional \emph{LeNet-5}~\cite{lenet} which was designed to recognize hand-written digits, the VGG19~\cite{simonyan_very_deep2014} classifier with 16 convolutional layers, and the Xception~\cite{xception} classifier with 36 convolutional layers. 
%
%What is remarkable is that even though current machine learning models do not resemble the complexity of its biological pendant, they inherently are able to generalize but only after learning thousands of examples. 
%Neural networks trained on one specific task can be used to perform detection or segmentation of, seemingly, unrelated objects with relatively minor retraining~\cite{transfer_learning}. 
%
%
%We think that the biological inspiration of modern convolutional neural networks yields the evaluation of principles of human perception with computers.
%
%
%MLP, LeNet, VGG, ImageNet, XCeption, ResNets etc and work from THomas Serre
%
%
%- https://link.springer.com/chapter/10.1007/978-3-642-14600-8_46
%
%- https://github.com/tidyverse/ggplot2/wiki/Recommended-Reading
%
%- last year's VADL workshop: https://vadl2017.github.io/

>>>>

\section{Related Work}

\noindent{\textbf{Graphical Perception.}} Cleveland and McGill~\cite{cleveland_mcgill,cleveland1985graphical} coined the phrase \emph{graphical perception} \change{to describe ``the visual decoding of information encoded on graphs''.} To understand encodings in visualizations, they define \emph{elementary perceptual tasks} as mental-visual stimuli, and rank their perceptual difficulty \change{to humans}. From these definitions, the authors perform the \emph{position-angle} experiment to compare bar charts and pie charts, and the \emph{position-length} experiment where users judge relations between encoded quantities in grouped and divided bar charts. The authors then use this to redesign a statistical map via \emph{bars and framed rectangles} and Weber's law \cite{harrison2014_webers_law_rank}, using the proportional relation between an initial distribution density and perceivable change.

Heer and Bostock later reproduced the Cleveland-McGill experiments via crowd-sourcing on Mechanical Turk~\cite{HeerBostock2010}, with similar results. Harrison et al.~repeated the Cleveland-McGill experiments while observing viewer emotional states, again with similar results~\cite{harrison2013influencing}. \change{Talbot et al.~delve deeper into the impact of specific bar chart design variations on human prediction \cite{Talbot2014}}. While we focus on Cleveland and McGill's work due to its repeated reproduction, many works investigate human perception to visual encoding~\cite{mackinlay1988applying,bertin1967semiologie,treisman1988feature,carpendale2003considering,wilkinson2006grammar,widgor_perception2007,munzner2015visualization}.

\change{
Cleveland and McGill's definition of graphical perception judiciously excludes a human subject and a visual decoding method, leaving the door open for machine perception such as with CNNs. Their quality standard is only that the machine must decode \emph{information}, which leaves narrow and well-defined applications like our experiments approachable. However, machine graphical perception must meet human breadth in capability to be generally useful in a human world.
}
%

@@

%
%\textbf{Comparing Visual Encodings.} Different visual encodings have advantages or disadvantages and the community does a great job comparing them. Higher-level comparisons include 2D versus 3D vector field and rendering studies~\cite{mckenzie_2d_3d,forsberg2009comparing_3d_vector,laidlaw_2d_vector,borkin2011arteries}, timeseries~\cite{herr2009timeseries} and scatterplots~\cite{tremmel1995visual,Wang_linegraph_vs_scatterplot}. Lower-level experiments target - besides others - open versus closed encodings~\cite{open_vs_closed_shapes}, and several evaluations of color space~\cite{ware1988color,rheingans1992color,Rogowitz2001_colormaps,kindlmann2002color}. While we investigate lower-level visual encodings in this work, we delay colorspace experiments for future work
%
%

@@

\\~\\
\noindent{\textbf{Visual-cortex-inspired Machine Learning.}} The human visual cortex allows us to recognize objects in the world seemingly without effort. This visual system is organized into layers, which inspired computational systems based on multilayer neural networks. Fukushima and Miyake developed the early Neocognitron quantitative model~\cite{fukushima1982neocognitron}, which ultimately led to the work of Hinton, Bengio, and LeCun~\cite{lecun2015deep} and today's GPU-powered deep neural networks. Such networks have been developed with many architectures \change{attempting to model properties useful to visual reasoning, like translation invariance or part hierarchies}. \change{Across these, the error behaviors of CNNs indicate that they process images in a different way to humans even though they are loosely biologically inspired~\cite{SzegedyICLR2014,Engstrom2017,Azulay2018}.} 
%
%For this paper, we compare a set of networks with different architectures and depths, plus networks with weights trained on natural images and on generated stimuli of the elementary perceptual reasoning tasks.
%
%Object detection is extremely difficult and therefore is especially impressive as light intensities can change by levels of magnitude and contrast between foreground and background is so often low. 
%In addition, the visual scene changes every time the human body or human eyes move. 
%This visual system exhibits a very noisy structure but because it is organized by layers it has inspired the mathematical theory of multilayer neural networks. 
%

@@

%
%The reported classification performance is superior to that of humans and the question in regards to their functionality opens an interesting research topic.
%Visual cortex inspired machine learning classifiers exist with many different architectures. We select the traditional \emph{LeNet-5}~\cite{lenet} which was designed to recognize hand-written digits, the VGG19~\cite{simonyan_very_deep2014} classifier with 16 convolutional layers, and the Xception~\cite{xception} classifier with 36 convolutional layers. 
%
%What is remarkable is that even though current machine learning models do not resemble the complexity of its biological pendant, they inherently are able to generalize but only after learning thousands of examples. 
%Neural networks trained on one specific task can be used to perform detection or segmentation of, seemingly, unrelated objects with relatively minor retraining~\cite{transfer_learning}. 
%
%
%We think that the biological inspiration of modern convolutional neural networks yields the evaluation of principles of human perception with computers.
%
%
%MLP, LeNet, VGG, ImageNet, XCeption, ResNets etc and work from THomas Serre
%
%
%- https://link.springer.com/chapter/10.1007/978-3-642-14600-8_46
%
%- https://github.com/tidyverse/ggplot2/wiki/Recommended-Reading
%
%- last year's VADL workshop: https://vadl2017.github.io/
%
%
\\~\\
\noindent{\textbf{Computational Visualization Analysis.}} Pineo et al.~create a computational model of early human vision based on neural networks, then optimize flow visualizations for comprehension~\cite{Pineo2012_computational_perception}. Their simulations show that visualization perception triggers neural activity in higher-level areas of cognition, which the authors suspect is supported by low-level neurons performing elementary perceptional tasks. Other work tries to parse infographics by finding higher-level saliency models~\cite{bylinskii2016should,bylinskii2017understanding,bylinskii2017predimportance}, \change{or tries to parse text and key visual elements from visualizations using both classic and deep-learning-based computer vision~\cite{Savva2011-revision,kembhavi2016diagram,zoya_text_visual_tags,maneesh_deconstructing_d3,Poco2017-reverse-engineering-vis} (to name but a few)}.

%Poco et al.~recover visual encodings from chart images by \cite{Poco2017-reverse-engineering-vis}
%
%Savva et al.~similarly recover chart elements by \cite{Savva2011-revision}.

%R. A. Al-Zaidy and C. L. Giles. Automatic extraction of datafrombarcharts. InProceedings of the 8th International Conference on Knowledge Capture, page 30. ACM, 2015. 2

\change{
The field of visual question answering (VQA) has begun to combine image and text feature vectors for visualizations. Kafle et al.~extend stacked attention networks with dynamic encodings to support different bar chart designs, plus present a benchmark dataset~\cite{kafle2018dvqa}. However, the ability of CNNs to solve visual relation problems like VQA has been called into question~\cite{not_so_clevr}. Kahou et al.~attempt to answer visual questions across five chart types. However, the authors state that the task poses ``a significant machine learning challenge'' with no tested system able to meet human-level performance~\cite{Kahou2018}.
}

\change{Our work takes a step back. None of these works investigate the building blocks of visualization: elementary perceptual tasks such as position, length, and angle estimation. To produce useful visualization analyzers, we must be able to computationally predict human responses in these tasks. We investigate whether this is possible with CNNs.}

%Feed forward neural networks - visual relations strain
